Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-08-28 paper.tex
Some work on the experiments section. Included a table with running times for queries with a cache hit.

\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.
%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
%computation accessing multiple data points in a kernel around the target location.}

@@

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

-- measure time taken to perform the thresholding computation for an entire time-step on a cold cache

-- measure time taken to evaluate the query after warming up the cache and polluting it afterward

The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

We observer nearly a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.4 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
the large variance that is seen in the running time of the queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.6 times faster when compared to
running a single thread.
\kk{Should we have scale-out experiments as well?}

Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieve from the
same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
I/O only with a single thread.

@@

- Lays the ground work for the creation of a landmark database, which could store the location of the highest vorticity regions in the dataset.

>>>>

\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{multirow}
\usepackage{array}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

%\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
%and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
%computation accessing multiple data points in a kernel around the target location.}

The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
The space required to cache the maximum number of points per time-step including the index space and database overhead is $\sim$40MB. 
Therefore for a dataset containing 
1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a derived field for the entire dataset will
require $\sim$5TB.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|c|c|} \hline
\multirow{2}{0.8in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{2-4}
 & No cache & \parbox[t]{0.7in}{With cache (miss)} & \parbox[t]{0.7in}{With cache (hit)} \\ \hline
4247 & 97.1 & 100.2 & 0.5\\ \hline
86580 & 113.7 & 115.9 & 1.2\\ \hline
909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
time that can be observed is due to the larger result set that has to be returned to the user.
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

Table \ref{tab:cache_effectiveness} shows the effectiveness of the cache in the case of a cache hit. It includes the data from Fig. \ref{fig:cache_overhead}
for the cases of execution in the absence of a cache and for cache misses. In the last column we show the running time of queries for which there is a 
cache hit. To simulate this scenario we first warm up the cache by submitting the same set of threshold queries of the vorticity field as before. We then submit
several more unrelated queries with different time-steps and threshold values in order to pollute the cache. Finally, we issue the original set of queries and 
measure their running times. As we can see from these results the introduction of an application-aware cache for the results of threshold queries yields over
an order of magnitude improvement for queries that hit the cache even for queries that have a low threshold and large result sets. Let us focus on the query
with low threshold, which returns $\sim$900,000 points. Given that the number of points that can be returned from a threshold query is limited to 1,000,000
it is likely that all subsequent queries to this time-step will result in a cache hit as their threshold is likely to be equal or higher than the cached one.

The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

We observer nearly a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.4 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.6 times faster when compared to
running a single thread.
\kk{Should we have scale-out experiments as well?}

Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieved from the
same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
I/O only with a single thread.

@@

- Lays the ground work for the creation of a landmark database, which could store the locations of the highest vorticity regions in the dataset.
